---
title: "Ch8Solutions"
author: "Jim Albert"
date: "1/10/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

#### Exercise 1: Career Trajectory of Willie Mays

(a) Use the get_stats function to extract the hitting data for Willie Mays for all of his seasons in his career.

Read in get_stats function.

```{r}
get_stats <- function(player.id) {
  Batting %>% 
    filter(playerID == player.id) %>%
    inner_join(Master, by = "playerID") %>%
    mutate(birthyear = ifelse(birthMonth >= 7,
                              birthYear + 1, birthYear),
           Age = yearID - birthyear,
           SF = ifelse(is.na(SF) == TRUE, 0, SF),
           SLG = (H - X2B - X3B - HR +
                    2 * X2B + 3 * X3B + 4 * HR) / AB,
           OBP = (H + BB + HBP) / (AB + BB + HBP + SF),
           OPS = SLG + OBP) %>%
    select(Age, SLG, OBP, OPS)
}
```

```{r}
library(tidyverse)
library(Lahman)
Master %>% filter(nameFirst == "Willie",
                  nameLast == "Mays") %>% 
  select(playerID) %>% pull() -> mays_id
mays_stats <- get_stats(mays_id)
```

(b) Construct a scatterplot of Mays' OPS season values against his age.

```{r}
ggplot(mays_stats, aes(Age, OPS)) +
  geom_point()
```

(c) Fit a quadratic function to Mays' career trajectory.  Based on this model, estimate Mays' peak age and his estimated largest OPS value based on the fit.

```{r}
fit <- lm(OPS ~ I(Age - 30) + I((Age  - 30) ^ 2), 
          data = mays_stats)
fit$coef
```

```{r}
A <- fit$coef[1]
B <- fit$coef[2]
C <- fit$coef[3]
(Peak.Age <- 30 - B / (2 * C))
(MAX <- A - B ^ 2 / 4 / C)
```


#### Exercise 2: Comparing Trajectories

(a) Using James' similarity score measure (function similar), find the five hitters with hitting statistics most similar to Willie Mays.

```{r}
vars <- c("G", "AB", "R", "H", "X2B", "X3B", 
          "HR", "RBI", "BB", "SO", "SB")
Batting %>% 
  group_by(playerID) %>% 
  summarize_at(vars, sum, na.rm = TRUE) -> C.totals
C.totals %>%
  mutate(AVG = H / AB,
         SLG = (H - X2B - X3B - HR + 2 * X2B +
                    3 * X3B + 4 * HR) / AB) ->
C.totals
Fielding %>% 
  group_by(playerID, POS) %>%
  summarize(Games = sum(G)) %>% 
  arrange(playerID, desc(Games)) %>% 
  filter(POS == first(POS)) -> Positions

C.totals %>%
  inner_join(Positions, by = "playerID") %>%
  mutate(Value.POS = case_when(
    POS == "C" ~ 240,
    POS == "SS" ~ 168,
    POS == "2B" ~ 132,
    POS == "3B" ~ 84,
    POS == "OF" ~ 48,
    POS == "1B" ~ 12, 
    TRUE ~ 0)) -> C.totals

similar <- function(p, number = 10) {
  C.totals %>% filter(playerID == p) -> P
  C.totals %>% 
    mutate(sim_score = 1000 -
             floor(abs(G - P$G) / 20) -
             floor(abs(AB - P$AB) / 75) -
             floor(abs(R - P$R) / 10) -
             floor(abs(H - P$H) / 15) -
             floor(abs(X2B - P$X2B) / 5) -
             floor(abs(X3B - P$X3B) / 4) -
             floor(abs(HR - P$HR) / 2) -
             floor(abs(RBI - P$RBI) / 10) -
             floor(abs(BB - P$BB) / 25) -
             floor(abs(SO - P$SO) / 150) -
             floor(abs(SB - P$SB) / 20) - 
             floor(abs(AVG - P$AVG) / 0.001) - 
             floor(abs(SLG - P$SLG) / 0.002) -
             abs(Value.POS - P$Value.POS)) %>%
    arrange(desc(sim_score)) %>% 
    head(number)
}
s_player <- similar(mays_id, number = 6) %>% 
  select(playerID) %>% pull()
s_player
```

(b) Fit quadratic functions to the (Age, OPS) data for Mays and the five similar hitters.  Display the six fitted trajectories on a single panel.

```{r}
get_stats2 <- function(playerid){
  d <- get_stats(playerid)
  d %>% mutate(Player = playerid)
}
map_df(s_player, get_stats2) -> alldata
ggplot(alldata, aes(Age, OPS)) +
  geom_point() +
  geom_smooth(method = "lm",
               formula = y ~ x + I(x^2),
              size = 1.5) +
      facet_wrap(~ Player, ncol = 3)
```


(c) Based on your graph, describe the differences between the six player trajectories.  Which player had the smallest peak age?

#### Exercise 3: Comparing Trajectories of the Career Hits Leaders

(a) Find the batters who have had at least 3200 career hits.

```{r}
Batting %>% 
  group_by(playerID) %>% 
  summarize(H = sum(H)) %>% 
  filter(H >= 3200) %>% 
  select(playerID) %>% pull() -> p3200
p3200
```

(b) Fit the quadratic functions to the (Age, AVG) data for this group of hitters, where AVG is the batting average.  Display the fitted trajectories on a single panel.

```{r}
get_stats_new <- function(player.id){
  Batting %>% 
    filter(playerID == player.id) %>%
    inner_join(Master, by = "playerID") %>%
    mutate(birthyear = ifelse(birthMonth >= 7,
                              birthYear + 1, birthYear),
           Age = yearID - birthyear,
           AVG = H / AB) %>%
    select(Age, AVG) %>% 
    mutate(Player = player.id)
}
map_df(p3200, get_stats_new) -> alldata
ggplot(alldata, aes(Age, AVG)) +
  geom_point() +
  geom_smooth(method = "lm",
               formula = y ~ x + I(x^2),
              size = 1.5) +
      facet_wrap(~ Player, ncol = 4)
```

(c) On the basis of your work, which player was the most consistent hitter on average?  Explain how you measured consistency on the basis of the fitted trajectory.

#### Exercise 4: Comparing Trajectories of Home Run Hitters

(a) Find the ten players in baseball history who have had the most career home runs.

(b) Fit the quadratic functions to the home run rates of the ten players, where HR.RATE = HR / AB.  Display the fitted trajectories on a single panel.

(c) On the basis of your work, which player had the highest estimated home run rate at his peak?  Which player among the ten had the smallest peak home run rate?

(d) Do any of the players have unusual career trajectory shapes?  Is there any possible explanation for these unusual shapes?

#### Exercise 5: Peak Ages in the History of Baseball

(a) Find all the players who entered baseball between 1940 and 1945 with at least 2000 career at-bats.

(b) Find all the players who entered baseball between 1970 and 1975 with at least 2000 career at-bats.

(c) By fitting quadratic functions to the (Age, OPS) data, estimate the peak ages for all players in parts (a) and (b).

(d) By comparing the peak ages of the 1940s players with the peak ages of the 1970s players, can you make
any conclusions about how the peak ages have changed in this 30-year period?







